---
title: "R Tools for Digital Twins - package dtRtools"
author: "Karline Soetaert and Stanley Nmor"
date: "15-Januari-2024"
output:
  html_document:
    df_print: paged
    toc: true
    number_sections: true
    toc_float: true
    code_folding: show
    theme: cosmo
abstract: >
 "The R-package dtRtools contains general tools (R-functions) that can be used for making ecological digital twins. It contains amongst others functions to read datafiles obtained from Dutch governmental organisations (which are in Dutch), or to read and prepare bathymetric data. Other functions allow fast interpolation of spatially-distinct timeseries to dense grids or to average timeseries." 
keywords:
  digital twin, wadden sea, marine, tools, "R"
bibliography: 
  vignettes.bib 
vignette: >
  %\VignetteIndexEntry{dtRtools: R tools for digital twins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(width = 120)
require(dtRtools)

knitr::write_bib(c(.packages()), 
                 file = "packages.bib") # adds all used packages to bib
knitr::write_bib(c("plot3D", "FME"), file = "packages.bib")
```

# Introduction

The package *dtRtools* contains tools that will be useful for making digital twins.

It contains several data sets:

-   *Marsdiep* contains the bathymetry of the marsdiep area in the Dutch waddensea.
-   *Shape* a shape file of the waddensea
-   *Wad_weather* contain wind speeds, light intensity, and radiation obtained from the KNMI (Dutch meteorological organisation) for the year 2021.
-   *Wad_watertemp_LR* contains low resolution surface water temperature from Waddensea stations in 2021 (Rijkswaterstaat data).
-   *Sediment*, sediment characteristics for the Waddensea


The package also contains several functions:

```{r}
ls("package:dtRtools")
```

-   Functions *read_KNMI*, *read_RWS* read data files obtained from Dutch governmental organisations; 
-   Function *read_bathymetry* reads bathymetric data in net cdf format, as obtained from EMODnet; .
-   Function *plot_bathymetry* plots bathymetric data
-   Functions *all_in_period*, *first_in_period*, *last_in_period*, *count_in_period* find first, last... occurrences of a variable meeting a certain target.
-   Function *average_timeseries* takes hourly, daily,... averages of time series
-   Function *dt_reshape*,  *dt_tolong*,  *dt_towide* are convenience functions to go from wide to long format
-   Functions *interpolate_xy* and *interpolate_xyt* provide fast interpolation of spatial or spatio-temporal data, where the input is sparse and the output dense
-   Function *average_xy* regrids high density spatial data by averaging them in a lower resolution regular grid


It also contains several small data files that are not exported, but used to demonstrate the used of the reading functions. They can be found in the packages extdata directory.

# Datasets

## Bathymetry

-   *Marsdiep* contains the bathymetry of the marsdiep area of the Dutch waddensea.

```{r}
meta(Marsdiep)
```

```{r, fig.width = 7, fig.height=5}
with(Marsdiep, 
   image2D(longitude, latitude, z=depth, asp=asp))
with(Marsdiep, invisible(lapply(contours, FUN=lines, col="white")))
mtext(outer=TRUE, side=3, "Datasets Waddensea and Marsdiep", cex=1.5)

```

## Abiotic conditions - weather data

data set *Wad_weather* contain wind speeds, light intensity, and radiation for the year 2021.

```{r}
head(Wad_weather)
meta(Wad_weather)
```

```{r, fig.width=8, fig.height=6}
par(mfrow=c(2,2))
Weather235 <- subset(Wad_weather, subset=station==235)
plot(Weather235, type="l")
```

## Abiotic conditions - water temperature

-   *Wad_watertemp_LR* contains water temperatures, at roughly monthly resoltuion, for several station in the Waddensea

```{r}
data(Wad_watertemp_LR)
head(Wad_watertemp_LR)
meta(Wad_watertemp_LR)
```

```{r, fig.width=8, fig.height=8}
plot(Wad_watertemp_LR, mfrow=c(3,4), type="l", las=1, ylab="dgC")
```

## Abiotic conditions - sediment data

```{r}
head(Sediment)
points2D(Sediment$longitude, Sediment$latitude, 
         colvar=Sediment$mdGrain, pch=16, asp=1.674109)
```

# Functions

# References

::: {#vignettes}
:::

# Acknowledgements

# Appendix

## List of all functions in the package


```{r,out.width=2000}
func_names <- capture.output(lsf.str("package:dtRtools"))
for (i in 1:length(func_names)) cat(func_names[i], "\n\n")
```
