---
title: "Chlorophyll distribution"
output: html_document
date: "2024-06-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The original formula

A function that generates a Chl profile, based on integrated concentration
in a layer of sediment with thickness zn. 
Based on the model from Jesus et al. 2006,

Journal of Experimental Marine Biology and Ecology 332 (2006) 60– 74

The formula of Jesus et al, 2006 represents the Chlorophyll amount accumulated at depth d $Chl_d$, as a function of the total amount of Chlorophyll in the layer with thickness $d_{max}$, $Chl_T$.

$$ Chl_d = Chl_T \cdot [ \frac{d}{d_{max}} + mud(1-\frac{d}{d_{max}}-e^{-2e\frac{d}{d_{max}}})]$$

At d=0, the integrated Chlorophyll equals 0, while at d=$d_{max}$ this equals $Chl_T$.

This is first rewritten as a weighted combination of the mud fraction and the remainder:

$$ IChl_d = \int_0^{d}f(z)dz=  Chl_T \cdot [ (1-mud)\cdot \frac{d}{d_{max}} + mud(1-e^{-2e\frac{d}{d_{max}}})]$$
This integral represents the integrated concentration of Chl up to a depth $d$, and f(z) is a function that expresses the Chl amount as a function of depth z. To find f(z), the right hand side is differentiated with respect to $d$.

Divided by $zn$, we get a function that describes the chlorophyll concentration at depth z, as a function of the mud content and the mean Chlorophyll concentration in a layer with thickness $d_max$, $Chl_m$:

$$ Chl_z = (1-mud) \cdot Chl_m + mud \cdot Chl_m \cdot (2 \cdot e \cdot e^{-2e\frac{z}{z_{max}}})]$$

```{r}
int.mud <- function(Chl  = 1,    # Integrated Chl concentration in layer 0-zn
                     zn   = 0.002,
                     pMud = 0.5){
  

  z     <- seq(0.5*zn/100, zn, by = zn/100)

  # integrated fraction
  Chl*(z/zn + pMud*(1-z/zn-exp(-2*exp(1)*z/zn)))
}
```


```{r}
pChl.mud <- function(Chl  = 1,    # Mean Chl concentration in layer 0-zn
                     zn   = 0.002,
                     pMud = 0.5){
  

  z     <- seq(0.5*zn/100, zn, by = zn/100)

  # Chl when 0 and 1 mudfraction
  Chl_0 <- Chl
  Chl_1 <- Chl*(2*exp(1)*exp(-2*exp(1)*z/zn)* (z < zn))
  Chlz  <- (1-pMud)*Chl_0 + pMud*Chl_1
  Chlz
}
```

Show that this is the same:

```{r}
x <- (0:99)/100
x.cum <-(0.5:99.5)/100
plot(int.mud(pMud=1)*100, x,ylab= "z/zn", xlab="Int conc", ylim=c(1,0))
lines(cumsum(pChl.mud(pMud=1)), x.cum)

points(int.mud(pMud=0.6)*100, x,col=2)
lines(cumsum(pChl.mud(pMud=0.6)), x.cum,col=2)

points(int.mud(pMud=0.3)*100, x,col = 2)
lines(cumsum(pChl.mud(pMud=0.3)),x.cum, col=3)

points(int.mud(pMud=0)*100, x,col=4)
lines(cumsum(pChl.mud(pMud=0)),x.cum, col=4)

```

This is how the actual concentration changes 

```{r}
x <- (0:99)/100
plot(pChl.mud(pMud=1), x, ylab= "z/zn", 
     xlab="concentration", ylim=c(1,0), type="l")

lines(pChl.mud(pMud=0.6), x, col=2)

lines(pChl.mud(pMud=0.3), x, col=3)

lines(pChl.mud(pMud=0), x, col=4)

```

```{r}
plot(diff(int.mud()))
lines(pChl.mud())
```

# Mean Chl concentration

```{r}
xm  <- 0.002
xn  <- 0.00025  # depth of the peak
xst <- 0.00005
x   <- seq(0, xm, len=10000)
y   <- dnorm(x, mean = xn, sd=xst) #- dnorm(0, mean = xn, sd=xst)
y   <- y/(pnorm(xm, mean = xn, sd=xst) - pnorm(0, mean = xn, sd=xst)) * xm
plot(x, y)
mean(y)
```
