---
title: "dtPP - Tools for primary production measurements - Photosynthesis-irradiance data"
author: "Karline Soetaert"
date: "02-October-2023"
output:
  pdf_document: default
  html_document: default
  word_document: 
abstract: 
 The R-package dtPP contains tools for estimating aquatic and terrestrial photosynthesis. 
 It comprises functions to convert raw photosynthesis data into production rates, 
 functions to fit photosynthesis-irradiance (PI) response curves, 
 and a function to calculate depth-integrated production. 
 Here we show the use of this package to estimate aquatic carrying capacity.
keywords:
  photosynthesis, algae, aquatic science, PI curve, "R"
bibliography: 
  vignettes.bib 
vignette: >
  %\VignetteIndexEntry{dtPP: Tools for estimating aquatic photosynthesis - Photosynthesis-irradiance data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
options(width = 120)
require(dtPP)
toBibtex(citation("dtTools"))
toBibtex(citation("dtWad"))
toBibtex(citation("dtPP"))
toBibtex(citation("plot3D"))
toBibtex(citation("FME"))
```

# Introduction


The R-package *dtPP* is one of a set of digital twin R-packages that deal with aquatic and terrestrial primary production 
  
Some of its functions can be used by aquatic scientists as a tool 
to estimate depth-integrated photosynthesis in aquatic environments. 
The same functions can also serve to estimate terrestrial primary production for instance in forests.

The package also contains several example datasets, for the Waddensea these are:

* *FRRFdata* with data as generated with the FRRF instrument (The acronym FRRF stands for Fast Repetition Rate Fluorometry). 
* *LABSTAFdata* for datasets generated with the labSTAF instrument.
* *Wad_biogeo_jetty* is a long-term dataset with biogeochemical data from the NIOZ jetty in the Waddensea.
* *Wad_chlor_RS* contains a spatial grid of chlorophyll, at monthly resolution, from the Waddensea and surrounding area in 2021.

Other datasets are made available from the dtWad digital twin package:

* *Marsdiep* and *Waddensea* contain the bathymetry of the marsdiep area and the Dutch waddensea respectively.
* *Grain* is a dataset with sediment characteristics in the Waddensea.
* *Weather2021* contain wind speeds, light intensity, and radiation for the year 2021 from a weather station near the Waddensea.
* *Height2021* and *Temp2021* contain water surface elevation (height) and surface water temperature from a few stations in 2021.


The package also contains several functions:

* Four mathematical models that describe PI responses: The functions *fWebb* @Webb1974,
  *fJP* @JassbyPlatt1976, *fPG* @PlattGalegosHarrison1980 and 
  *fEP* @EilersPeeters1988.
* Four related functions: *fWebb.I, fJP.I, fPG.I, fEP.I*, are the standardized versions 
  (respective to I).
* Functions to fit the PI responses to these model or to newly defined mathematical models.
These functions make use of the FME @FME package.
* A function to estimate depth-integrated photosynthesis.

# Datasets

## FRRF data sets

There are two sets of data with primary productivity data

* *FRRFdata* is one data set, generated with the FRRF instrument.
* *LABSTAFdata* is a data set generated with a labSTAF instrument.

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,2), oma=c(0,0,2,0))
with(FRRFdata, plot(E, JVPII, pch=18, cex=2,
                    ylab="mmol photons/m3/hour", main="JVPII"))
with(FRRFdata, plot(E, Fo, pch=18, cex=2))   
with(FRRFdata, plot(E, Fm, pch=18, cex=2)) 
with(FRRFdata, plot(E, Fo/Fm, pch=18, cex=2)) 
mtext(outer=TRUE, side=3, "Dataset FRRFdata", cex=1.5)
```

```{r, fig.height=8, fig.width=8}
par(mfrow=c(2,2), las=1, oma=c(0,0,2,0))
with(LABSTAFdata, plot(E, JVPII, col=factor(file), pch=18, cex=2))
with(LABSTAFdata, plot(E, Fo,    col=factor(file), pch=18, cex=2))   
with(LABSTAFdata, plot(E, Fm,    col=factor(file), pch=18, cex=2)) 
with( LABSTAFdata, plot(E, Fo/Fm, col=factor(file), pch=18, cex=2)) 
mtext(outer=TRUE, side=3, "Dataset LABSTAFdata", cex=1.5)
```

## Biogeochemical data

* *Wad_biogeo-jetty* is a long-term dataset with biogeochemical data from the NIOZ jetty in the waddensea.
* *Wad_chlor_RS* contains a spatial grid, monthly resolution, from the Waddensea and surrounding area in 2021 obtained from remote sensing.

The units and meaning of the variables in the Jetty data (Wad_biogeo_jetty) is stored in its attributes:

```{r}
attributes(Wad_biogeo_jetty)$variables
```

```{r, fig.height=8, fig.width=8}
par(mfrow=c(3, 3), las=1, oma=c(0,0,2,0))
with(Wad_biogeo_jetty, plot(datetime, Temperature, type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, Secchi, type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, Chl,  type="l", lwd=1)) 
with(Wad_biogeo_jetty, plot(datetime, NO3,  type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, NO2,  type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, NH4,  type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, Si,   type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, Ntot, type="l", lwd=1))
with(Wad_biogeo_jetty, plot(datetime, DON,  type="l", lwd=1))
mtext(outer=TRUE, side=3, "Dataset Jetty", cex=1.5)
```

The remote sensing data are rather crude:

```{r, fig.width=10, fig.height=10}
mf <- par(mfrow=c(3,3), las=1, oma=c(0,0,2,0))

image2D(x=Wad_chlor_RS$longitude, 
        y=Wad_chlor_RS$latitude, 
        z=Wad_chlor_RS$chlorophyll[,,2:10], 
        asp=Wad_chlor_RS$asp, 
        log="c", clim=c(0.1, 50),
        clab="Chl, mg/m3", xlab="dgE", ylab="dgN",
        main=month.abb[2:10])
mtext(outer=TRUE, side=3, "Dataset Wad_chlor_RS", cex=1.5)
```

# Four mathematical descriptions of photosynthesis - response curves

Four functions to fit PI curves are included: fWebb, Webb (1974); fJP, Jassby and Platt (1976); fPG, Platt, Galegos and Harrison (1980); and fEP, Eilers and Peeters (1988);

```{r, fig.cap="The four PI models included in the package"}
I <- 0:1500
plot(fWebb(I), type="l", lwd=2, 
     main="Photosynthesis functions",
     ylab="mgC/m3/hr", xlab="uEinst/m2/s")
lines(fJP(I),  col=2, lwd=2)
lines(fPG(I),  col=3, lwd=2)
lines(fEP(I),  col=4, lwd=2)
legend("bottomright", legend = c("fWebb", "fJP", "fPG", "fEP"), col=1:4, lwd=2)
```

# Fitting photosynthesis-irrandiance data sets

Fitting one PI response data set can be done with function *fitPI* whose 
arguments are:

```{r}
args(fitPI)
``` 

To demonstrate its use, we first create the vectors with the data:

```{r}
I <- seq(0, 100, length.out = 11)
response <- c(0, 1.6, 1.85, 1.95, 1.99, 2.0, 1.98, 1.95, 1.92, 1.88, 1.82)
```
They are fitted to each of the PI models, using function *fitPI*.

```{r}
FitWebb <- fitPI("Webb", I = I, response = response)
FitPG   <- fitPI("PG",   I = I, response = response)
FitJP   <- fitPI("JP",   I = I, response = response)
FitEP   <- fitPI("EP",   I = I, response = response)
```

The fitting algorithm used (*modFit* from the *FME* package) returns
a list of class *modFit* and *FRRF*, that can be interrogated with the usual 
functions *summary*, *coef*, *deviance* and so on. 

We can look at some results with *summary*

```{r}
summary(FitWebb)
summary(FitPG)
```

The plotting function shows the fit with the data:

```{r}
plot(FitWebb)
```

The SSR of the various models is:

```{r}
c(deviance(FitWebb), deviance(FitPG), deviance(FitJP), deviance(FitEP))
```

Another way to visualise the correspondence of models versus data is to 
generate predictions and plot those:

```{r}
par(mfrow = c(1, 1))
plot(I, response, ylim = c(0, 2))

# I for prediction
PARpred <- seq(0, 120, by = 1)

lines(predict(FitWebb, PARpred), lwd = 2)
lines(predict(FitPG, PARpred), col = "red", lwd = 2)
lines(predict(FitJP, PARpred), col = "blue", lwd = 2)
lines(predict(FitEP, PARpred), col = "green", lwd = 2)
legend("bottomright", c("Webb", "PG", "JP", "EP"), 
   col = c("black", "red", "blue", "green"), lty = 1, lwd = 2)

# use plot to look at single fits
par(mfrow = c(2,2))
plot(FitWebb, main = "Webb")
plot(FitPG, main = "Platt-Galegos")
plot(FitJP, main = "Jassby-Platt")
plot(FitEP, main = "Eilers-Peeters")
```

# Fitting a newly defined model to one data set

## A simple model

It is possible to define a different model and fit that to a PI dataset.

Let's implement a simple Monod function; this should have as input arguments
irradiance at which to calculate the reponse, and the parameter vector, in 
that order; it should return a list with I and response vector:

```{r}
Monod <- function(I, p) {
  y <- p[1]*I / (I + p[2])
  list(x = I, y = y)
}
```

The data are similar as above:

```{r}
I <- seq(0, 100, length.out = 11)
response <- c(0, 1.6, 1.85, 1.95, 1.99, 2.0, 1.98, 1.95, 1.92, 1.88, 1.82)
```

To fit the new model, we need to pass initial guesses of the parameters (*pini*).

```{r}
FitMonod <- fitPI(Monod, I = I, response = response, pini = c(Rm = 2, ks = 25))
```

We look at some results

```{r}
summary(FitMonod)
coef(FitMonod)
```

The short version of visualising the model fit is to use function *plot* on the fitting object:

```{r, fig.cap="Visualising model fit, using function plot"}
plot(FitMonod)
```

# A more complex model

We now define a more complex Monod model where the response also depends on temperature,
according to a Q10 formulation.

There are thus 3 parameters in this model: the maximal photosynthesis rate (Rm), 
the half-saturation light intensity (ks) and the temperature coefficient (Q10)

In addition to the irrandance and parameter vector, this model also has
as input argument the temperature (either one value or a vector of length = length of I):

```{r}
MonodT <- function(I, p, T) {
  
  y <- p[1]*I / (I + p[2]) * p[3]**((T-10)/10)
  list(x = I, y = y)
}
```

The input data now also include the temperature - assume here that temperature
dropped during the experiment.

```{r}
I <- seq(0, 100, length.out = 11)
T <- seq(15, 10, length.out = 11) 
response <- c(0, 1.6, 1.85, 1.95, 1.99, 2.0, 1.98, 1.95, 1.92, 1.88, 1.82)
```

The call to *fitPI* now needs to pass also the temperatures; this is 
done in the *modarg* (model arguments) argument, a list with named objects.

```{r}
FitMonodT <- fitPI(MonodT, I = I, response = response, 
  pini = c(Rm = 2, ks = 25, Q10 = 2), modargs = list(T = T))
```

We look at some results

```{r}
summary(FitMonodT)
coef(FitMonodT)
```

To apply *plot*, we need also to pass the input temperature. 
in the *plot* function, the response is predicted on I values that form a sequence of 
100 values, from the minimal to maximal I. Hence, the additional temperature
vector passed via *modargs* should also be of length 100: 

```{r}
par(mfrow = c(1, 1))
plot(FitMonodT, modargs = list(T = seq(15, 10, length.out = 100)), 
  main = "Monod*f(Temp)")
```

# Fit models to multiple data sets at once

Sometimes several samples should be fitted at once.
An example of such a data set is included in the package and called *LABSTAFdata*

```{r}
data(LABSTAFdata)
head(LABSTAFdata)
```

For demonstration purposes, we select the first 5 sets of data;
as we will select column *JVPII* for fitting, we already remove
those records where this measurement is *NA*:

```{r}
 set  <- subset(LABSTAFdata, subset = (!is.na(JVPII)))
 tail(set)
```

We now fit these all at once using the Eilers-Peeters model:

```{r}
 mFitEP <- fitmultiPI("EP", id = set$file, 
                      I=set$E, 
                      response=set$JVPII, 
                      normalized = TRUE)
```

The summary gives the best parameter values with their standard errors and
fitting probabilities for each data set (ids is the index in PIdat):

```{r}
 summary(mFitEP)  # look at summary
```

We can also extract just the coefficients:

```{r}
 coef(mFitEP)     # extract parameters
```

We can plot all fits at once:

```{r, fig.height=8, fig.width=8}
par(mfrow = c(1, 1))
plot(mFitEP)
```

Alternatively, we can predict the model fits and plot them in one plot:

```{r, fig.cap="Visualising multiple model fits, using function predict and matplot"}
 Par <- 1:1500 
 predEP <- predict (mFitEP, I = Par)
 head(predEP)

 par (mfrow = c(1, 1))   
 matplot(Par, predEP, type = "l", lwd = 2, col = 1:10, lty = 1)
 legend("topright", legend = 1:10, lwd = 2, col = 1:10, lty = 1)
 points(set[,c("E", "JVPII")], col = (1:10)[set[,"file"]], pch = ".", cex = 4)
```

#  Fit newly defined model to multiple data sets at once

We now use the Monod function again, and fit it to the multiple data sets

```{r}
 Monod <- function(I, p) {
  y <- p[1]*I / (I + p[2])
  list(x = I, y = y)
 }

 mFitMonod <- fitmultiPI(Monod, set$file, set$E, set$JVPII, normalized = TRUE, 
  pini = c(pm = 1, ks = 1))
 summary(mFitMonod)
```

# References

<div id="refs"></div>


# Acknowledgements

The initial version of some of these functions were created in cooperation with 
Jacco kromkamp and Gregg Silsbe.

