---
title: "Estimating integrated photosynthesis using R-package dtPP"
author: "Karline Soetaert, Stanley Nmor, Qing, Leon"
date: "04-October-2023"
output:
  pdf_document: default
  html_document: default
  word_document: 
abstract: We use functions from the R-package dtPP to estimate depth-integrated photosynthesis in the aquatic environment for a number of typical cases. We start with a simple example where the meteorological conditions and photosynthesis parameters are constant. We then impose time-varying light intensities, fluctuating water levels, and time-varying PI parameters.  
keywords:
  photosynthesis, algae, aquatic science, PI curve, "R"
bibliography: 
  vignettes.bib 
vignette: >
  %\VignetteIndexEntry{Estimating integrated photosynthesis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
options(width = 120)
require(dtPP)
toBibtex(citation("dtTools"))
toBibtex(citation("dtWad"))
toBibtex(citation("dtPP"))
toBibtex(citation("plot3D"))
toBibtex(citation("FME"))
```

# A simple example

In the simplest example, all inputs (depth, light intensity, extinction coefficient, and photosynthetic parameters) are constant.

Integrated photosynthesis are estimated for a water depth of 10 m

```{r}
P1 <- integratedPP(
  zmax    = 10,      # depth of water column, [m]
  times   = 0,       # time for which to estimate PP, one value
  
  # Eilers-Peeters photosynthesis parameters
  PI.par=c(alpha = 0.05, # (mg C m-3 h-1) * (uE m-2 s-1)-1
           eopt  = 350,  # (uE m-2 s-1)
           ps    = 6),   # mg C m-3 h-1
  
  It.data = 100,     # light intensity at surface, [uEinst/m2/s]
  kz      = 1)       # extinction coefficient, [/m]
```

The function *integratedPP* returns a list with several items: 

* The timeseries of integrated values, 
* vertical profiles of time-averaged values, and 
* time-depth (2D) data of light and volumetric photosynthesis.

The function *meta* shows what these variables mean, and their units.

```{r}
names(P1)
meta(P1)
```


```{r}
P1$ts
names(P1$profile)
plot(P1, which=c("PP_v", "Iz_I0"))
```

Integrated photosynthesis are estimated for a water depth of 10 m.

```{r}
P2 <- integratedPP(
  zmax    = 10,      # depth of water column, [m]
  times   = 0:100,       # time for which to estimate PP, one value
  
  # Eilers-Peeters photosynthesis parameters
  PI.par=c(alpha = 0.05, # (mg C m-3 h-1) * (uE m-2 s-1)-1
           eopt  = 350,  # (uE m-2 s-1)
           ps    = 6),   # mg C m-3 h-1
  
  # light intensity at surface, [uEinst/m2/s], timeseries
  It.data = data.frame(time=0:100, 
                       value=seq(100,2000, length.out=101)),     
  kz      = 1,        # extinction coefficient, [/m]
  unit    = c(mass = "mg C", length = "m", time = "h", light = "uE m-2 s-1")
)
```


# A complex example

This example uses measured irradiance data (*It.data*), fluctuating water levels (*Ht.data*), 
and time-varying Photosynthesis-irradiance parameters (*PI.par*). 
The integrated photosynthesis is estimated for a 4 -day period. 

The times for which we want to run the model is input in *POSIXct* data format.

```{r}
time <- seq(from = as.POSIXct("2021-08-01 01:00:00"), 
            to   = as.POSIXct("2021-08-05 01:00:00"), 
            by   = 60)  # output every minute
```

The light intensity has been obtained from the KNMI; the data are stored in a data.frame called *Wad_weather* in the *dtPP* package.

The data from station Nr 235 are extracted from the data. 

The light values are called "radiation" in the dataset; they are in J/m2/s. 
Data are converted to uEinst/m2/s, by multiplying with 4.6.
It is assumed that 50% of this light is PAR.

```{r}
st235   <- subset(Wad_weather, 
                  subset = station==235)
It.data <- data.frame(datetime = st235$datetime, 
                      PAR      = 0.5* 4.6 * st235$radiation)
```

Water height varies in time (~tide), according to a sine wave, with a periodicity of 12.4 hours.

```{r}
t.period    <- 12.4*3600   # tidal period in seconds
t.amplitude <- 1           # tidal amplitude in meters

wH       <- t.amplitude * sin(2*pi*as.double(time)/t.period)
Ht.data  <- data.frame(t      = time, 
                       height = wH)
``` 

The photosynthesis-irradiance parameters alfa and ps are Chl-specific values. 

As the chlorophyll changes over time, so do these two PI-parameters. 

The used PI parameters are calculated by multiplying the CHl-specific values with chlorophyll.

```{r}
Chl <- data.frame(time = as.POSIXct(c("2021-08-01 01:00:00", 
               "2021-08-02 01:00:00", "2021-08-03 01:00:00", 
               "2021-08-04 01:00:00", "2021-08-05 01:00:00")),
               conc = c(1, 2, 4, 3, 1))
```

The photosynthesis-irradiance curves were fitted with the Eilers-Peeters model.

```{r}
# Eilers-Peeters photosynthesis parameters
pars = list(alpha = 0.05, # (mg C m-3 h-1) * (uE m-2 s-1)-1
            eopt  = 350,  # (uE m-2 s-1)
            ps    = 6)    # mg C m-3 h-1

PI.par <-   data.frame(
    time  = Chl$time,    
    alpha = pars$alpha*Chl$conc,   
    eopt  = 180, 
    ps    = pars$ps   *Chl$conc)

```

Estimate depth integrated production

```{r}
unit  <- c(mass   = "mg C", 
           length = "m", 
           time   = "h", 
           light  = "uE m-2 s-1")

system.time(
  PP <- integratedPP(zmax    = 5,       # water depth
                   times   = time,    # output time
                   It.data = It.data, # light intensity timeseries
                   kz      = 1.5,     # extinction coefficient
                   PI.par  = PI.par,
                   Ht.data = Ht.data, # water height (surface elevation)
                   unit    = unit)
)

PP2 <- integratedPP(zmax    = 0.5,       # water depth
                   times   = time,    # output time
                   It.data = It.data, # light intensity timeseries
                   kz      = 1.5,     # extinction coefficient
                   PI.par  = PI.par,
                   Ht.data = Ht.data, # water height (surface elevation)
                   unit    = unit)
```

```{r, fig.width=8, fig.height=8}
par(mfrow=c(3,2))
plot(PP, PP2, which=c("PP", "TotalDepth", "Isurf", "Ibot"), lwd=2, mfrow=NULL)
lines(PP2$ts$times, PP2$ts$PP, col="red")
with(PP$ts, plot(times, alpha, 
                 xlab=" time", ylab="..", main="alpha", 
                 lwd=2, type="l", las=1))
with(PP$ts, plot(times, ps,
                 xlab=" time", ylab="..", main="ps", 
                 lwd=2, type="l", las=1))
```

