---
title: "Estimating integrated photosynthesis in the Wadden Sea using R-package dtPP"
author: "Karline Soetaert"
date: "24-December-2023"
output:
  pdf_document: default
  html_document: default
  word_document: 
abstract: 
keywords:
  photosynthesis, algae, Wadden sea, PI curve, "R"
bibliography: 
  vignettes.bib 
vignette: >
  %\VignetteIndexEntry{Estimating integrated photosynthesis in the Wadden Sea using R-package dtPP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
options(width = 120)
require(dtPP)
toBibtex(citation("dtLife"))
toBibtex(citation("dtWad"))
toBibtex(citation("dtPP"))
toBibtex(citation("plot3D"))
toBibtex(citation("FME"))
```

# A complex example

This example combines measured irradiance data, fluctuating water levels, and time-varying PI-parameters and extinction coefficients with remote sensed chlorophyll data to estimate photosynthesis. 

The integrated photosynthesis is estimated for a 4 -day period. 
The times for which we want to run the model is input in POSIXct data format.

```{r}

time <- seq(from = as.POSIXct("2021-01-01 00:00:00"), 
            to   = as.POSIXct("2021-03-31 23:30:00"), 
            by   = 60*30)  # output 30 minutes
```

The light intensity has been obtained from the KNMI; they are stored in a data.frame called *Wad_weather*.
The data from station Nr 235 are closest to the measurment site. 
The light values are called "radiation" in the dataset; they are in J/m2/s.
Assume that 50% of light is PAR.

```{r}
st235   <- subset(Wad_weather, 
                  subset = 
                    station == 235                                 &
                    datetime >= as.POSIXct("2020-12-31 23:59:59")  &
                    datetime <= as.POSIXct("2021-04-01 01:00:10")  )
It.data <- st235[, c("datetime", "radiation")]
It.data$PAR <- It.data$radiation * 0.5
It.data$radiation <- NULL

```

## Chlorophyll

Monthly remote sensing data for 2021 are stored in a 3D array (longitude x latitude x time).

The data are recast in long format, retaining only values within the Waddensea and removing absent (NA) values.

```{r}
# convert 3D array to long format
ChlLong <- with(Wad_chlor_RS, 
                expand.grid(nr        = 1:nrow(Wad_chlor_RS$datetime), 
                            longitude = longitude, 
                            latitude  = latitude))

ChlLong$date.from   <- Wad_chlor_RS$datetime[ChlLong$nr, 1]
ChlLong$date.to     <- Wad_chlor_RS$datetime[ChlLong$nr, 2]
ChlLong$chlorophyll <- as.vector(Wad_chlor_RS$chlorophyll)

# mask data only in WaddenSea
ChlLong$mask <- mask_xy(ChlLong[, c("longitude", "latitude")], 
                        shape = Wad_shape)$mask
ii      <- which(is.na(ChlLong$mask))
ChlLong <- ChlLong[-ii,]
ChlLong <- na.omit(ChlLong)
ChlLong <- subset(ChlLong, subset = 
                    date.from <= as.POSIXct("2021-04-01") &
                    longitude < 6.)
```

# Water height and temperature 

For water height and water temperature  we use the data from station DENHDR and DENHDVSGR

```{r}  
Ht.data  <- Wad_waterheight_HR[ , c("datetime", "DENHDR")]
Ht.data  <- subset(Ht.data, 
                  subset =                                         
                    datetime >= as.POSIXct("2020-12-31 23:59:59")  &
                    datetime <= as.POSIXct("2021-04-01 01:00:10"))

Tt.data  <- Wad_watertemp_HR[ , c("datetime", "DENHDVSGR")]
Tt.data  <- subset(Tt.data, 
                  subset =                                         
                    datetime >= as.POSIXct("2020-12-31 23:59:59")  &
                    datetime <= as.POSIXct("2021-04-01 01:00:10"))
``` 


Photosynthesis parameters change over time. Assume that alpha and ps are Chl-specific values, so multiplying with chlorophyll gives the values to be used.

# PI parameters

 values for alpha, eopt and ps are not available from the Waddensea

The values in Brinkman and Jacobs (2023) are used instead (approximately) 

 Brinkman & Jacobs, 2023. Gross pelagic primary production in the Ems-Dollard estuary,
 Journal of Sea Research 192 (2023) 102362

 alpha and ps are Chl-specific values -> they need to be multiplied with Chl
 at each time step

 eopt and ps depend on temperature

```{r}
Tmon <- average_timeseries(Tt.data, avgOver="mon", 
                           value = "DENHDVSGR")
```

Stations and their depth

```{r}
xy_Stats <- unique(ChlLong[,c("longitude", "latitude")])
Chl_Stats <- map_xy(input.x  = Wad_depth$longitude, 
                    input.y  = Wad_depth$latitude, 
                    input.2D = Wad_depth$depth,
                    output.xy = xy_Stats)
```

Estimate depth integrated production

```{r}
temp <- as.data.frame(approx(Tt.data, xout=time))  


PPall <- NULL
for (i in 1:nrow(Chl_Stats)){
  Chl_data <- subset(ChlLong, subset=longitude == Chl_Stats$longitude[i]  & 
                                     latitude  == Chl_Stats$latitude[i]   )
  Chl_data <- Chl_data[, c("date.from", "chlorophyll")]
  Chl      <- as.data.frame(approx(Chl_data, xout=time))

# time-variable PI parameters
# steepness of the light curve, mgC/mgChl/(uEinst/m2/s)
#  alpha <- 0.049  

# optimal light intensity (in uEinst/m2/s - original values in W/m2)
#  eopt  <- 4*(150 + 15*ChlRS$Temperature)

# assumed temperature dependence for maximal gross production
  ps    <- 13*1.06^(temp$y-20)    # mgC/mgChl/h

  PI.par <- data.frame(time  = time,
                       alpha = 0.049*Chl$y,
                       eopt  = 4*(150 + 15*temp$y),
                       ps    = ps*Chl$y)
 zmean <- Chl_Stats$z[i]
PP <- integratedPP(zn      = zmean,   # water depth
                   times   = time,    # output time
                   It.data = It.data, # light intensity timeseries
                   kz      = 0.3,     # extinction coefficient
                   PI.par  = PI.par,
                   Ht.data = Ht.data, # water height (surface elevation)
                   avgOver = "day")

PPall <- cbind(PPall, PP$ts$PP)
}
colnames(PPall) <- paste("ST",format(Chl_Stats$longitude, digits=3), "E_", 
                              format(Chl_Stats$latitude,  digits=3), "N", sep="")
PPall <- data.frame(datetime = PP$ts$time, PPall)
```

```{r, fig.width=8, fig.height=8}
matplot(PPall$datetime, PPall[,-1], type="l", 
        lty=1, ylab="mgC/m2/h", las=1, lwd=2,
        main="PP wadden")

legend("topleft", legend=names(PPall)[-1], col=1:10, lty=1)
```

