---
title: "Digital Twin of the Wadden Sea - package dtWad"
author: "Karline Soetaert"
date: "02-October-2023"
output:
  pdf_document: default
  html_document: default
  word_document: 
abstract: 
 The R-package dtWad contains general tools (R-functions), and data sets that can be used for the ecological digital twins of the Wadden Sea. It contains functions to read datafiles obtained from Dutch governmental organisations (which are in Dutch!), or to read and prepare bathymetric file. Other useful functions are to convert the spatial coordinates used in the Netherlands to more convenient latitude-longitude coordinates, or to interpolate spatially-distinct timeseries to other spatial points. 
keywords:
  digital twin, wadden sea, marine, tools, "R"
bibliography: 
  vignettes.bib 
vignette: >
  %\VignetteIndexEntry{dtWad: tools and data for Wadden Sea digital twins}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup, include=FALSE}
options(width = 120)
require(dtWad)
toBibtex(citation("dtLife"))
toBibtex(citation("dtWad"))
toBibtex(citation("plot3D"))
toBibtex(citation("FME"))
```

# Introduction

The general context of the digital twin (not too long)

The package dtWad contains tools and datasets that will be useful for other twins that deal with specific topics.

It contains several data sets:

* *Waddpth* contains the bathymetry of the Dutch waddensea 
* *Wad_sediment* is a dataset with sediment characteristics in the Waddensea.
* *Wad_weather* contain wind speeds, light intensity, and radiation for the year 2021.
* *Wad_waterheight_HR* and*Wad_watertemp_HR* contains water surface elevation (height) and surface water temperature from a few stations in 2021.

The package also contains several functions:

* Functions to read data files obtained from Dutch governmental organisations: *readKNMI*, *readRWS*
* Functions to read bathymetric data in net cdf format, and that is obtained from EMODnet: *readBathymetry*.


It also contains several small data files that are not exported, but used to demonstrate the used of the reading functions. They can be found in the packages extdata directory.


# Datasets

## Bathymetry

* *Marsdiep* and *Wad_depth* contain the bathymetry of the marsdiep area and the Dutch  respectively.

```{r, fig.width = 7, fig.height=10}
par(mfrow=c(2,1), oma=c(0,0,2,0), las=1)
with(Wad_depth, {
   image2D(longitude, latitude, z=depth, asp=asp)
  invisible(lapply(contours, FUN=lines))
  })

with(Marsdiep, 
   image2D(longitude, latitude, z=depth, asp=asp))
with(Marsdiep, invisible(lapply(contours, FUN=lines, col="white")))
mtext(outer=TRUE, side=3, "Datasets Wad_depth and Marsdiep", cex=1.5)

```

## Abiotic conditions

* *Wad_sediment* is a dataset with sediment characteristics in the Waddensea.
* *Weather2021* contain wind speeds, light intensity, and radiation for the year 2021.
* *Wad_waterheight_2021_HR* and *Wad_watertemp_2021_HR* contain water surface elevation (height) and surface water temperature from a few stations in 2021.

```{r, fig.width=8, fig.height=6}
par(mfrow=c(2,2))

asp <- 1/cos((mean(Wad_sediment$latitude)*pi)/180)
with(Wad_sediment, points2D(longitude, latitude, colvar=mdGrain, 
       log="c", clab="um", main="Median grain size",
       pch=".", las=1, cex=2, asp=asp))         
with(Wad_sediment, points2D(longitude, latitude, colvar=silt, 
       log="c", clab="", main="percent silt",
       pch=".", las=1, cex=2, asp=asp))         
with(Wad_sediment, points2D(longitude, latitude, colvar=sand, 
       log="", clab="", main="percent sand",
       pch=".", las=1, cex=2, asp=asp)) 
with(Wad_sediment, points2D(longitude, latitude, colvar=sand, 
       log="", clab="", main="percent sand",
       xlim=c(4.5, 5), ylim=c(52.8, 53.4),
       pch=".", las=1, cex=2, asp=asp))        
```


# References

<div id="refs"></div>


# Acknowledgements

